---
title: "GLLVMs A-Å"
subtitle: "Some inputs into performing a full analysis using GLLVMs"
institute: "Department of Mathematical Sciences, NTNU"
author: "Audun Rugstad, Ph.D. candidate"
date: "April 2025"
format: 
  beamer:
    slide_level: 2
    section-titles: true
    #aspectratio: 32
    navigation: frame
    theme: Dresden
    colortheme: spruce
    fontfamily: opensans
    #fonttheme: smallcaps
header-includes:
  - \usepackage[dvipsnames]{xcolor}
  #- \setbeamercolor{frametitle}{fg=black}
  #- \setbeamertemplate{itemize items}[default]
  #- \useoutertheme{miniframes}
  #- \setbeamercolor{enumerate items}{fg=ForestGreen}
  - \setbeamercolor{itemize item}{fg=ForestGreen}
  - \setbeamercolor{itemize items}{fg=ForestGreen}
  - \setbeamercolor{itemize subitem}{fg=ForestGreen}
  - \setbeamercolor{itemize subsubitem}{fg=ForestGreen}
bibliography: references.bib
---

# Intro

## Outline of the session

Five-step approach:

1.  Formulate your question in terms of a statistical model

2.  Data exploration and preparation

3.  Model fitting

4.  Model checking

5.  Making inferences

## Outline of the session

We'll go through the full analysis of an ecological dataset

**Example:** @dou2022: *Influence of environmental variables on macroinvertebrate community structure in Lianhuan Lake*

# Formulating the question

## Formulating the question

Formulate your research question in terms of a statistical analysis

First question: **What is the goal of the analysis?**

-   **Prediction:** Find the model that best predict future observations

-   **Explanation:** Investigate causal relationships between explanatory variables and response(s)

## Prediction

::: {.callout-tip icon="false"}
## Prediction

-   Find the model that best predict future values of one/more response variables (i.e. species distribution)

-   Variable selection based on optimizing for prediction

-   I.e. forward/backward selection using AIC, Root mean square error etc.
:::

**Example:** A model that predicts future distributions of wood beetles under different climate change scenarios

## Explanation

Two types:

\small

::: {.callout-tip icon="false"}
## Confirmatory

-   Test a **specific, clear hypothesis** with one/a few models

-   Predictor variables selected based on *a priori* knowledge

-   Ideally **no variable selection** and pre-registration
:::

. . .

::: {.callout-tip icon="false"}
## Exploratory

-   Use model(s) to **explore possible** causal relationships

-   Should only be used to generate hypotheses

-   Avoid automatic model selection
:::

\normalsize

## Flowchart

```{r fig.width=4, fig.height=2.5}
#| echo: false
library(ggflowchart)
data <- tibble::tibble(from = c("Model", "Model", "Explanatory", "Explanatory"), to = c("Predictive", "Explanatory", "Confirmatory", "Exploratory"))
ggflowchart(data)
```

## How does this apply to community ecology

Exploratory "throwing everything at the wall" approaches perhaps most common

JSDMs more associated with prediction than other methods?

"True" confirmatory analyses are rare

-   The line between the three can often be blurry

## How does this apply to community ecology

::: {.callout-tip icon="false"}
## How can GLLVMs help?

-   More streamlined and focused analysis

-   Ideally, a single model of the community that is as complete as possible

-   Can then be used to make inferences about different causal relationships etc.
:::

## How does this apply to community ecology

**Q:** What type of analysis does a GLLVM with only **unconstrained** latent variables fit into?

. . .

**A:** Primarily exploratory, generating hypothesis about factors that structure the community.

*But:* if the goal is to test specific hypothesis about species co-occurrences, could potentially be used more confirmatory

## How does this apply to community ecology

**Q:** What type of analysis does a GLLVM with **constrained** latent variables fit into?

. . .

**A:** All three, depending on approach?

## Example: Invertebrates in Lihuan

@dou2022: Macroinvertebrates in 13 lakes in Lihuan, CN:

From the intro: *The community composition of macro-invertebrate assemblages and their relationships with environmental variables were investigated*

What type of modeling approach makes sense here?

. . .

**A:** Seems very clearly **exploratory** (which variables are potential drivers of community change?)

## Example: Invertebrates in Lihuan

What should the model look like?

. . .

A **concurrent ordination** might be a good fit

-   Allows for assessing the relative effect of many predictor variables at once, on a small number of latent variables

-   Also allows us to account for potential residual variation in the L.V.s (i.e. variation not explained by the predictors)

## Example 2

Haven't yet found a good dataset/study that is not exploratory in the same way:(

# Data preparation

## Data exploration and preparation

Important to always **visually inspect** key properties of the data before modeling

-   Ensure the data meets the assumptions of the model

-   Act as a "sanity check" for modeling output

Standardize and scale variables as needed

As with most other models, we recommend "common sense" practices and guidelines such as @zuur2010

## Data exploration and preparation

Some factors that can mess up a GLLVM:

-   Predictor variables that are too co-linear

-   Several species with very little information on abundances (i.e. zero-inflation)

-   Sites with very little information on abundances

-   Highly unbalanced sampling

## Model philosophy

The underlying philosophy of GLLVMs is:

-   Data is ideally gathered with a spesific model or analysis in mind

-   If not the case: tailor a model to the data properties

-    The model should account for different properties of the data, as far as possible

T**he model should suit the data**, not the other way around.

## Example: Lake invertebrates

```{r}
#| include: false
library(ggplot2)
library(mefa)
library(ggalluvial)
library(dplyr)
library(tidyr)
library(magrittr)
library(tibble)
library(stringr)
library(GGally)
library(gllvm)
#library(hellinger)
library(ggpubr)
set.seed(1)
```

\footnotesize

```{r}
#| eval: true
#| echo: true

# load data
lake_sp <- read.csv("../data/lake_inverts_sp.csv", sep = "\t")
lake_env <- read.csv("../data/lake_inverts_env.csv", sep = "\t")

# clean (according to paper)
lake_sp <- lake_sp |> select(-Site, -Group, -Season, -Lakes)
# reduce to actual counts!
lake_sp <- lake_sp / 16
```

## Example: Species

We can use a (Log-)abundance occupancy plot:

```{r}
#| fig-height: 4
# plot abundance-occupancey (function from package mefa)
aoplot(lake_sp)
```

**X** = Number of plots a species occupies, **Y** = average log-abundance

Can help identify outlier species + get a sense of the commonness of species in the dataset

## Species data

Table of total species abundances:

\footnotesize

```{r}
#| code-overflow: wrap
table(colSums(lake_sp))
```

\normalsize

9 species with 1-2 occurrences in the dataset, and three species with very high abundances

## Filtering data

**Suggestion:** remove species occurring in only one plot, and plots containing only one species (remove 8 species and 9 plots).

\footnotesize

```{r}
#| echo: true

# remove rare species
lake_sp_filt <- lake_sp |> 
  select_if(colSums(lake_sp !=0)>1) 

# remove rare sites
lake_sp_filt <- lake_sp_filt[rowSums(lake_sp_filt != 0)>1,]

# remove same sites in env. dataframe
lake_env_filt <- lake_env[row.names(lake_sp_filt),]

```

\normalsize

## Environmental variables

Correlation of all predictors (lake chemical attributes):

:::: columns
::: {.column width="100%"}
\tiny

```{r}
#| echo: false 
#| fig-height: 2
corrplot::corrplot(cor(lake_env_filt |> select(-Site, -Group, -Season, -Lakes)),
                   type="lower", method="pie", tl.cex =0.4)
```
:::
::::

\normalsize

## Environmental variables

Why is this important?

-   High co-linearity makes it hard for the model to converge (find estimates)

-   Also makes the parameter uncertainties larger

## Environmental variables

What can be done?

-   Omit predictors

    -   Need to consider properties of the system and analysis

-   "Model tricks"

    -   Regularization ("reigning in" wild estimates)

    -   Shrinkage (finding ways to make unimportant parameters -\> 0)

    -   Ex: Switching to random effects or putting constraints on the response distribution

## How high is too high?

Rule of thumb: no more than 0.8

Our max cor = 0.69

So we "just" standardize and scale to mean = 0 and variance = 1:

```{r}
#| echo: true
lake_env[,5:17] <- scale(lake_env[,5:17])
```

-   So all coefficient estimates are on the same magnitude

# Model setup

## Model setup

**Our suggestion:** Concurrent ordination

What to include in the model?

-   Our goal is to explore the impact of **environmental variables**

-   As such, we can maybe argue for ignoring the study design here (differences between season/lake assumed to be manifested through the water properties)

## Model setup

**Initial model setup:**

-   Two concurrent latent variables informed by the chemical properties of the water

-   Random row effects (to account for site variation in overall abundance)

-   Poisson distribution

-   *Could* include quadratic effects, but might be overloading the model

## Model setup

\footnotesize

```{r}
#| echo: true
#| eval: false
#| code-overflow: wrap

mod_lakes <- gllvm(lake_sp_filt, # species dataframe
                   X = lake_env_filt, #envrionemtnal dataframe
                   lv.formula = ~ WT + DO + PH + CON + CODMn + 
                         TP + TN + NH3.N + NO3.N + NO2.N + Chla + 
                         SS + WD,
                   row.eff = "random", # random row effect
                   family = "poisson", # poisson distribution
                   num.lv.c = 2, #number of concurrent LVs
                   n.init = 20) # run the model a few times 
                                # to ensure convergence
```

```{r}
# load the actual model
load("mod_pois_1.Rdata")
```

## Model setup

A few (personal) tips for the `gllvm()` function:

-   `n.init`: How many starting iterations you run

    -   Depending on how long one run takes, maybe 5-50. Tells you when it has converged.

-   `trace = TRUE`: tells you when each model run specified in `n.init` is complete (makes it less frustrating to fit)

-   `sd.errors = FALSE/TRUE`: when diagnosing/comparing models, you can turn off estimation of standard errors to make each model fitting faster (and fit them retroactively using the `se.gllvm()` function)

\normalsize

# Model checking

## Model checking with GLLVMs

What to look at to determine whether the model is "good"?

-   **Diagnostic plots** (`plot()` function) for model assumptions

    -   Most important!

-   AIC/BIC to compare model fit (e.g. number of LVs needed)

-   Goodness of fit-tests (g`llvm.goodnessOfFit()`)

-   Very small/large parameter estimates (particularly sp. coefficients) for convergence

## Model checking with GLLVMs

What to do to improve the fit?

-   Change response distribution

-   Run more iterations

-   Change starting values

-   Change from fixed to random effects

-   Change number of LVs

## Model checking: Example

We can look at diagnostic plots:

```{r}
#| layout-ncol: 3
plot(mod_lakes_pois)
```

We see some **fanning** and potential **overdispersion** in the residual and QQ plots.

## Model checking: Example

After re-running the model with `family = "negative.binomial"`:

```{r}
#| layout-ncol: 3

load("mod_nb_1.Rdata")

plot(mod_lakes_nb)
```

The residual plots indicate that the NB model meets the assumptions of the model better.

## Model checking: Example

The difference in **AICc** values between model 1 and 2 also indicates that the negative binomial model is better.

```{r}
#| echo: true
AICc(mod_lakes_pois, mod_lakes_nb)
```

## Model checking: Example

Looking at some **goodness-of-fit** measures to the original data, however...

```{r}
gof_pois <- goodnessOfFit(y=lake_sp_filt, object = mod_lakes_pois)
gof_nb <- goodnessOfFit(y=lake_sp_filt, object = mod_lakes_nb)
data.frame(Measure=c("cor", "RMSE", "MAE", "MARNE"),
           Poisson=as.numeric(gof_pois),
           NB=as.numeric(gof_nb))
```

. . .

**Conclusion:** Our N.B. model, while meeting the model assumptions (a valid model), is a pretty bad fit to our data

## What has happened?

If we look at the plots of the standard errors for the species loadings, we also see that some are very large, indicating a poor convergence of the model:

\footnotesize

```{r}
#| echo: true
#| fig-height: 4

plot(mod_lakes_nb$sd$theta) # theta = species coefficients
```

\normalsize

## What has happened?

If we look at the residual

\footnotesize

```{r}
sum_mod <- summary(mod_lakes_nb)
sum_mod$sigma.lv
```

\normalsize

## Improving the NB model

We try two things:

-   Specify the LV-coefficients as **random effects**, drawn from a distribution unique for each coefficient

-   Tell the model to estimate a common *dispersion* parameter for the negative binomial distribution for each species

Both could help in steering the model away from "extreme" estimates and lessen overfitting (=regularizing)

## Improving the NB model

\footnotesize

```{r}
#| echo: true
#| eval: false
#| code-overflow: wrap

mod_lakes_nb_2 <- gllvm(lake_sp_filt, X = lake_env_filt,
                   lv.formula = ~ WT + DO + PH + CON + CODMn + 
                         TP + TN + NH3.N + NO3.N + NO2.N + Chla + 
                         SS + WD,
                   row.eff = "random",
                   family = "negative.binomial",
                   ## LV coefficients are random effects:
                   randomB = "P", 
                   ## Only one dispersion parameter estimated:
                   disp.formula = rep(1, ncol(lake_sp_filt)),
                   num.lv.c = 2,
                   n.init = 20)
```

```{r}
load("mod_nb_2.Rdata")
```

## Checking the new model

```{r}
#| layout-ncol: 3

plot(mod_lakes_nb_2)
```

## Checking the new model

::::: columns
::: {.column width="50%"}
```{r}
plot(mod_lakes_nb_2$sd$theta)
```
:::

::: {.column width="50%"}
\tiny

```{r}
data.frame(Measure=c("cor", "RMSE", "MAE", "MARNE"),
           #Poisson=as.numeric(gof_pois),
           NB1=as.numeric(gof_nb),
           NB2= as.numeric(goodnessOfFit(y=lake_sp_filt, object = mod_lakes_nb_2)))
```
:::
:::::

## 

# Making inferences

## Making inferences with GLLVMs

What you look at depends on your question.

Typical visualizations:

-   Model summary

-   Biplots and triplots

-   Variance explained

-   Coefficient plots (caterpillar plots)

## Model summary

```{r}
sum_mod <- summary(mod_lakes_nb_2)
```

Two most important parts for us:

**`Residual standard deviation of LVs`**:

\tiny

```{r}
sum_mod$sigma.lv
```

\normalsize

Very low, implies our predictors explain ≈ all the variance in the latent variables

We *could* go back and fit the model again without residual LV variation (as a "classic" constrained ordination)

## Model summary

**`Coefficients LV predictors:`**

```{r}
sum_mod$REbcovs
```

The most importation predictors seem to be `WT` (water temperature), `PH` (water PH), and `Chla` (chlorophyll A content)

Relatively similar to the conclusion in @dou2022.

## Model summary

We can also look at the estimates for the random row effects:

```{r}
plot(sum_mod$`Row intercepts`)
```

They are very small, and could potentially also be excluded to refine the model.

## Biplot and triplot

```{r}
ordiplot(mod_lakes_nb_2)
```

## Biplot and triplot

```{r}
ordiplot(mod_lakes_nb_2, biplot=T)
```

## Variance explained

\tiny

```{r}
#| echo: true

varPartitioning(mod_lakes_nb_2)

```

\normalsize

## Coefficient plots

```{r}
#| layout-ncol: 2
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("Chla"))
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("NH3.N"))
```

## Coefficient plots

```{r}
#| layout-ncol: 2
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("WT"))
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("PH"))
```

## Takeaway so far?

. . .

-   While the model seems well behaved and reasonably fit to the data, estimates for the effect of predictors on the species are **very** high

-   Suggests it it not very useful for explanatory purpose (only very weak evidence)

## Comparison to fully constrained model

To get a sense of the problem, we could compare it to the slightly more "shaved" model that the parameter estimates seem to suggest:

\tiny

```{r}
#| echo: true
#| eval: false
mod_lakes_nb_3 <- gllvm(lake_sp_filt, X = lake_env_filt,
                   lv.formula = ~ WT + DO + PH + CON + CODMn + 
                         TP + TN + NH3.N + NO3.N + NO2.N + Chla + 
                         SS + WD,
                   ## Remove random row effects:
                   #row.eff = "random", 
                   family = "negative.binomial",
                   ## LV coefficients are random effects:
                   randomB = "P", 
                   ## Only one dispersion parameter estimated:
                   disp.formula = rep(1, ncol(lake_sp_filt)),
                   ## Fully constrained (reduced rank) coefs:
                   num.RR = 2,
                   n.init = 20)
```

```{r}
load("mod_nb_3.Rdata")
```

\normalsize

## Model checking

```{r}
#| layout-ncol: 3
plot(mod_lakes_nb_3)
```

## Biplot (model 2)

```{r}
ordiplot(mod_lakes_nb_3)
```

## Variance explained (model 2)

\tiny

```{r}
#| echo: true

varPartitioning(mod_lakes_nb_3)

```

\normalsize

## What has changed?

. . .

The effect of Chlorophyll and NH3 content has **disappeared**!

## Coefficient plots (model 2)

```{r}
#| layout-ncol: 2
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("WT"))
randomCoefplot(mod_lakes_nb_2, which.Xcoef = c("PH"))
```

Still extremely high uncertainty

## Conclusion

. . .

Both models can be argued to be a reasonable, but suggest different drivers of species diversity.

-   Still, some indication that PH and water temperature are important factors to look into (stood out in both models)

-   Large uncertainties like this often a feature of negative binomial models

If the goal was prediction, we could have done automatic model selection or similar to (probably) get a simpler model with clearer estimates

## Conclusion

**Highlights** the need to not trust model outputs blindly

The tools showed here can be useful in figuring out what is going on.

Next time: Try to collect more data in a way that could fit a Poisson model?

-   Some unclearness about the actual sampling effort and pooling of samples in this study

-   Unbalanced study design can have had an influence(more sampling in alkaline lakes than

## Conclusion

In other words:

-   We can say both **that** more data is needed, and something about **why** more data is needed:)

# References

## References
